Helixer: Cross-species Gene Annotation Of Large Eukaryotic Genomes Using Deep Learning

### 2020-06-04 paper/main.tex
hope it's actually ready to send around

\title[short Title]{Helixer: Cross-species Gene Annotation Of Large Eucaryotic Genomes Using Deep Learning}
\author[Sample \textit{et~al}.]{Felix Stiehler\,$^{\text{\sfb 1}}$, Marvin Steinborn\,$^{\text{\sfb 1}}$, Stephan Scholz, Daniela Dey, Andreas P.M. Weber\,$^{\text{\sfb 1,}*}$, Alisandra K. Denton\,$^{\text{\sfb 1}}$}
\address{$^{\text{\sf 1}}$Institue for Plant Biochemistry, Heinrich-Heine-University, Dusseldorf, D-40225, Germany}

@@

\abstract{
\textbf{Motivation:} Current state-of-the-art tools for the \textit{de novo} annotation of 
genes in eukaryotic genomes have to be specifically fitted for each species and still often 
produce annotations that can be improved much further. The fundamental algorithmic 
architecture for these tools has remained largely unchanged for about two decades, limiting
learning capabilities. Here, we set out to improve the cross-species annotation of
genes from DNA sequence alone with the help of deep learning. The goal is to eliminate 
the dependency on a closely related gene model while also improving the predictive
quality in general with a fundamentally new architecture.\\
\textbf{Results:} We present Helixer, a framework for the development and usage of 
a cross-species deep learning model that improves significantly on performance and
generalizability when compared to more traditional methods. We evaluate our 
approach by building a single model for the annotation of 186 animal genomes as 
well as one for 51 plants. Our predictions are shown to be much less sensitive to
the length of the genome than those of a current state-of-the-art tool. We also 
present two novel post-processing techniques that each worked to further strengthen 
our annotations and show in-depth results of an RNA-Seq based comparison of our 
predictions.\\
\textbf{Availability:} The source code of this work is available at 
\url{https://github.com/weberlab-hhu/Helixer} under the GNU General 
Public License v3.0.\\
\textbf{Contact:} \href{andreas.weber@hhu.de}{andreas.weber@hhu.de}\\  
\textbf{Supplementary information:} Supplementary data are available at 
\textit{Bioinformatics} online.
}

@@

For the purpose of gene annotation, RNNs have already shown promising results. 
\citep{choudharypredicting} carries out preliminary explorations on the potential 
of BLSTMs for cross species gene prediction and trains his model on human genes to 
test it later on two more species. DeepAnnotator \citep{amin2018deepannotator} uses 
bidirectional LSTMs (BLSTM) for gene finding in prokaryotes. Gene prediction in 
prokaryotes is considered more amenable than in eukaryotes, as genes in prokaryotes 
are proportionately more frequent in the genome, feature simpler control structures 
and do not use splicing \citep{wang2004brief}. DanQ \citep{quang2016danq} proposes 
the use of a BLSTM after a convolutional neural net (CNN) to find detailed motifs in 
the human genome. DeePromotor \citep{oubounyt2019deepromoter} trains a similar 
architecture for the recognition of promotor regions. Recently, several groups 
\citep{jaganathan2019predicting, wang2019splicefinder} successfully used CNNs to 
find splicing sites.

In this work we present Helixer, a novel prototype software for training and 
utilizing a general purpose DNN for the ab-initio cross-species gene annotation of 
large eukaryotic genomes using only DNA sequence as input. Our model is trained to 
differentiate between 4 regions: Intergenic, Untranslated (UTR), Coding (CDS) and 
Intron. We demonstrate the effectiveness of this approach by training a single model 
for the annotation of a large set of genomes from the kingdoms metazoa and 
viridiplantae, respectively, which we will call {\it animals} and {\it plants} from 
now on. We worked with the full data of 192 animal genomes and 60 plant genomes. 
Both the ability to generalize across genomes as well as the scope of the evaluation 
greatly surpasses all other work in this field. The source code and all input data 
is publically available.

\begin{methods}
\section{Methods}
\subsection{Datasets}
The foundation of our work are 192 animal and 60 plant genomes. The data of each 
genome consists of the latest publically available genomic assembly in form of a 
FASTA file and the latest annotation in the GFF format. One genome for each species of animal 
in EnsemblMetazoa 45 (\cite{howe2020ensembl}; Table S\ref{suptab:downloads_animals}) 
as well as all non-embargoed plant species from the JGI Phytosome 13 database 
(\cite{goodstein2012phytozome}; Table S\ref{suptab:downloads_plants}); The exact 
genomes are listed in the aformentioned tables.
Data was downloaded on the 15/10/2019 and 29/03/2019 for animals and plants respectively. 
Both data groups were used seperately during training as well as during evaluation.

@@

For the selection of training genomes we had to balance multiple conflicting 
tradeoffs. On the one hand, we want to train with as much data as possible, but 
on the other
not all data has sufficient quality to enable a powerful generalization. It is 
also very desirable to have diverse training genomes with a broad phylogenetic spread, 
a variety of genome sizes and average gene lengths as the model is increasingly unlikely to generalize 
well beyond the borders given by the training data. However, it also may be difficult 
for the model learn and generalize if the genomic patterns inside the data are 
too different from each other. Practically, it was important for us to limit the size of
our training genomes set to be small enough that enough that we could get experimental 
results within a couple of days and thereby be able to test many different data and 
model configurations.

@@

We call the genomes that were ultimatily not used during training from the original 
split the {\it validation genomes}. While we did split off a group of test genomes 
at the beginning of our work, we ended up only reporting on the generalization 
performance on the combined set of validation and test genomes as the validation 
genomes are by far the largest set. We also did not find a noticible difference in 
performance between those two groups.

@@

During data generation, we query for the transcript with the longest protein of each gene and 
disregard FASTA sequences that have no functional annotation as it is ambiguous
whether such sequences contain no genes or were simply not annotated in the reference. GeenuFF checks the 
genomic annotations for potential errors during import and is able to mark those 
areas. We use this information to effectively blank out those bases during training 
by using the sample weights described in Table \ref{tab:arrays}. The vast majority 
of masked bases lie in the intergenic region as the most prevelant error is a 
missing UTR and GeenuFF marks a potentially large intergenic region for it. Table 
\ref{tab:statistics_all} shows statistics about the masking.

For the training itself, we divide each continous genomic sequence into 20,000 bp 
long subsequences, for which one-hot vectors of the base pairs and annotations 
are generated and respectively used as input and label for the neural network 
together with the sample weights. We append zero padding if the subsequences  
are shorter than 20,000 bp. 
If a subsequence is fully marked as erroneous, we exclude it from all analyses.

\begin{table}[!htb]
\processtable{Data group statistics for all data\label{tab:statistics_all}} {
\begin{tabular}{@{}lll@{}}
\toprule & Animals & Plants\\
\midrule
Average genome size in Gbp& 2.936 (+- 1.562) & 0.787 (+- 0.995) \\
Average gene length & 31,223 (+- 13,974)& 3,368 (+- 1,510)\\
Geenuff error rate & 0.311 (+- 0.129) & 0.351 (+- 0.249) \\
Fraction of class Intergenic  & 0.777 (+- 0.042) & 0.799 (+- 0.089) \\
Fraction of class UTR & 0.006 (+- 0.006) & 0.017 (+- 0.016) \\
Fraction of class CDS & 0.016 (+- 0.014) & 0.085 (+- 0.071) \\
Fraction of class Intron & 0.201 (+- 0.035) & 0.099 (+- 0.057) \\
\botrule
\end{tabular}}{{\it Note:} Values are averages of the individual values of each 
genome in a data group. All statistics except the average gene lengths exclude
FASTA sequences without a gene and any 20,000 bp subsequences that were masked as completely 
erroneous. Each strand of DNA was counted seperately. The gene length was 
determined by the length of the pre-mRNA of the longest protein at each
loci. Brackets show the standard deviation.}
\end{table}

@@

\subsection{Model Architecture and Training}
\label{sec:model}
We use a 4-layer deep stacked BLSTM network with 256 units per layer and layer 
normalization \citep{ba2016layer} between each BLSTM layer to produce the 
predictions in the form of a base pair wise classification. The model consists of 
circa 5.4 million parameters and was implemented with the deep learning library 
Keras (https://keras.io) on top of TensorFlow \citep{abadi2016tensorflow}. We 
tested multiple different architectures before arriving at this model configuration, 
including convolutional neural networks (CNN) and hybrid architectures. We also use 
class weigths as the class frequencies are both unbalanced and 
vary greatly between animals and plants. 
For example, the tradeoff between the number of intergenic and intronic bases is 
quite different in both groups (see Table \ref{tab:statistics_all}).

@@

\subsection{Evaluation against independent RNAseq data}
\label{sec:rnaseq}
The reference annotations were created with existing tools and largely with a pipeline incorporating
{\it de novo} gene predictions with RNAseq and homology data. As the references, like any
data, are expected to contain errors we chose to use RNAseq data for an independent evaluation.
In plants there is the additional concern that the references may share biases with
AUGUSTUS or HMMs as AUGUSTUS was used as the {\it de novo} gene caller for the reference of
many plant species % (e.g. {\it Arabidopsis lyrata}, \cite{rawat2015improving}; 
%{\it Asparagus officinalis}, 
%\cite{harkess2017asparagus}; and {\it Chenopodium quinoa}, \cite{jarvis2017genome})  
and other HMM-based tools for many more. %(e.g. {\it Citrus clementina}, \cite{wu2018genomics}; and
%{\it Capsella grandiflora}, \cite{slotte2013capsella}). 
The Ensemble animal dataset by and large
used the Ensembl annotation pipeline, which includes no HMM-based tool.

@@

% todo, add versions to everything below
For each of these six species the following search was performed on Sequence Read Archive
`(("<species name>"[Organism] OR <species name>[All Fields]) AND stranded[All Fields]) 
AND ("biomol rna"[Properties] AND "library layout paired"[Properties])'. In so far as 
more than 50 samples were identified and every Nth sample was selected so that under 50
samples were chosen for further processing (see Table S\ref{suptab:rnaseq_samples}).
Each sample was quality controlled with FastQC (www.bioinformatics.babraham.ac.uk/projects/fastqc/), before and after trimming
with Trimmomatic \citep{bolger2014trimmomatic}, and then mapped to the genome with Hisat2 \citep{kim2019graph}. 
Samtools \citep{li2009sequence} was used for the sorting and indexing that was necessary both for
RNAseq mapped metric collection with PicardTools (http://broadinstitute.github.io/picard/) and custom downstream 
processing below. MultiQC \citep{ewels2016multiqc} was used to summarize the results. The above was 
automated and the results visualized with the code here (... make available and link to RNAseq stuff).
The samples were filtered based on many different metrics. Relatively high mapping rates, high mapping to
exonic regions relative to other regions, low 3' bias, normal FastQC, Trimmomatic stats and a confirmation
to be stranded with 2nd strand as sense strand were all taken into account. If more remained, seven
of the high quality samples were selected randomly.

Finally, the selected and mapped RNAseq samples were merged and quantified to
get the coverage (number of reads matching; i.e. cigar =, M or X),
and spliced coverage (number of reads with gap or splice; i.e. cigar N or D)
for every base pair in the genomes as implemented in the "training\_rnaseq.py"
script.

@@

Our models, however, tend to perform less well for the very smallest and largest 
genomes or species that are phylogenetically the furthest away from our training 
genomes. This is the case for both animals and plants and is visualized by the 
Figures S\ref{supfig:phylogeny_f1_animals} and S\ref{supfig:phylogeny_f1_plants}. 
We do not, for example, consistently predict very well on the 
algae nor on the non-avian reptiles. While one of the the algae were included 
in our training genomes it accounted for a tiny proportion of the total training 
data and the non-avian reptiles were not included at all.

@@

\begin{figure*}[!h]
\label{fig:cov_example_main}
\centerline{\includegraphics[width=\textwidth]{images/cov_examples/cov_example_main}}
\caption{Four example helixer predictions in the context of RNAseq data, the reference,
and AUGUSTUS' prediction for {\it M. esculenta}. The examples were chosen so
that {\bf a)} Helixer had high accuracy against the reference and the reference
was supported by the RNAseq data, {\bf b)} Helixer had high accuracy but the 
reference was not supported by the RNAseq data, {\bf c)} Helixer had low accuracy
and the reference was supported by the RNAseq data, and {\bf d)} helixer had low
accuracy but the reference was not supported by RNAseq data. Plotability was
a major secondary consideration. Each subplot shows from top to
bottom i) the natural log of the coverage (``cov", solid) and spliced coverage 
(``sc", dotted) + 1, ii) the reference annotation in matrix form, iii) 
AUGUSTUS' predictions in matrix form, and iv) Helixer's predictions. The reference
and AUGUSTUS have either 0 (white) or 1 (black) for each base pair and category, while
Helixer emits a probability from 0-1 represented via gray-scale. ``Ntrn" stand
for intron, and ``IG" stands for intergenic.
}
\end{figure*}

As an independent evaluation and to get a better qualitative understanding of
performance we compared Helixer predictions, AUGUSTUS predictions and the
reference to RNAseq coverage data. RNAseq was processed for three each of plants
and animals chosen because they had varied performance quality relative to AUGUSTUS. 
RNAseq coverage provides support for an 
exonic annotation (UTR or CDS), spliced coverages provides support for an 
annotation of intron, and neither coverage nor spliced coverage is expected 
for intergenic annotations. Looking at selected subsequences (Figure \ref{fig:cov_example_main}) % and supplemental figures.
we identified cases where RNAseq supported a) both the annotations of the reference 
and Helixer, b) neither, c) the reference, but not Helixer,
and d) Helixer, but not the reference.

Helixer models do not yet have postprocessing
to make finalized single predictions, but instead output basewise probabilities.
We see that the model exibits higher uncertainty around transitions between annotation 
classes, for instance between UTR and CDS, or more dramatically between UTR and 
intergenic even where Helixer predictions closely match the reference and RNAseq data (a). 
Helixer's uncertainty around transitions from UTR to intergenic regions may relate to a
fundamentally harder problem (there is no conserved motif at the site as is observed
for splice sites and start/stop codons), lack of a precise one 
base pair biological site \citep{carninci2006genome,hon2013quantification}
or noise in the reference, which we observed relative to the RNAseq data 
(Figures S\ref{supfig:UTRs_animals}-\ref{supfig:UTRs_plants}). % supplemental UTR figures, 1/species

Helixer models also sometimes showed uncertainty for larger regions.
In some cases, where Helixer did not receive RNAseq support for its highest probability
annotation, it assigned a low but non-trivial probability to the RNAseq-supported 
exon/intron pattern (Figures \ref{fig:cov_example_main}b-c, S\ref{supfig:cov_example_02}c). However, in the extreme,
there are cases where the Helixer model exibits substantial indecision or confusion 
and shifts gradually between classes with no single class receiving a high probability 
for extended stretches (Figures S\ref{supfig:cov_example_02}c-d). 
Notably, in one of the examples (Figure S\ref{supfig:cov_example_02}d)
the RNAseq shows evidence of alternative splicing; 
and in another (Figure S\ref{supfig:cov_example_02}c) Helixer's
prediction falls between that of the reference and of AUGUSTUS.

@@

Coverage and Spliced coverage were broken down by the confusion matrix of both
Helixer vs AUGUSTUS as well as Helixer vs the reference for all genomes (Figure \ref{fig:average_vs_augustus}). 
Where both tools agreee on far left in the figure, coverage and spliced coverage closely matched expectations. 
Specifically for CDS:CDS and UTR:UTR most base pairs had some, and many
had moderate or high coverage. The same pattern was seen for intron:intron and spliced
coverage. Finally, intergenic:intergenic showed only a small fraction of base pairs with 
any of either coverage or spliced coverage. In all conflicts the amount of RNAseq support
fell between the cases where tools agreed, indicating that all options were at least
capable of finding weak-spots in the other annotations.

In cases of conflict, Helixer's predictions were substantially more consistent with
the RNAseq data than those of AUGUSTUS. Specifically, for base pairs where one tool predicted CDS and
the other intergenic or intron, there was more coverage when 
Helixer predicted CDS (Figure \ref{fig:average_vs_augustus}a). Similarly, for base pairs where one tool predicted intron and the other 
CDS or intergenic, there was more spliced coverage when Helixer predicted intron (Figure \ref{fig:average_vs_augustus}b). 
These patterns were consistent in direction but varied in magnitude in the individual species,
with the expception of {\it P. marinus} where Helixer and AUGUSTUS performed comparably
at differentiating between CDS and intron regions (Figure S\ref{supfig:each_sp_vs_augustus}).

RNAseq based comparison of Helixer and the reference was less clear cut.
Averaged across species, helixer's predictions received slightly more support
than the reference when differentiating CDS, UTR and introns from intergenic;
however helixer's predictions received slightly less support than the reference 
when differentiating CDS and UTR from introns (Figure S\ref{supfig:average_vs_ref}). Performance varies between individual
species, from {\it P. marinus}, where the reference receives more support
in every conflict, to {\it T. cacao}, where Helixer models receive equivalent
or more support in every conflict (Figure S\ref{supfig:each_sp_vs_ref}). Interestingly, these are the two species that
were selected because they had (relative to the usual difference) low Subgenic F1 compared to 
AUGUSTUS; for the former the RNAseq confirms relatively weak performance for Helixer, while the 
for the latter RNAseq rather indicates a sub-par reference.

@@

Finally, developent of a post-processing method
to go from base pair wise predictions to integrated predictions for whole 
transcripts at each loci could both further improve performance and would
greatly increase real world applications.

\section*{Funding}
This work has been supported by the Deutsche Forschungsgemeinschaft (DFG, German Research Foundation) 
under Germany's Excellence Strategy--EXC-2048/1--Project ID: 390686111.\vspace*{-12pt}

\section*{Acknowledgements}
We are grateful for the computational suppport and GPU access provided by
the de.NBI cloud, and by both the Centre for Information and Media Technology (ZIM)
and the Institute of Quantitative and Theoretical Biology at Heinrich Heine University.

>>>>

\title[short Title]{Helixer: Cross-species Gene Annotation Of Large Eukaryotic Genomes Using Deep Learning}
\author[Sample \textit{et~al}.]{Felix Stiehler\,$^{\text{\sfb 1}}$, Marvin Steinborn\,$^{\text{\sfb 1}}$, Stephan Scholz, Daniela Dey, Andreas P.M. Weber\,$^{\text{\sfb 1}}$, Alisandra K. Denton\,$^{\text{\sfb 1,}*}$}
\address{$^{\text{\sf 1}}$Institue for Plant Biochemistry, Heinrich-Heine-University, Dusseldorf, D-40225, Germany}

@@

\abstract{
\textbf{Motivation:} Current state-of-the-art tools for the \textit{de novo} annotation of 
genes in eukaryotic genomes have to be specifically fitted for each species and still often 
produce annotations that can be improved much further. The fundamental algorithmic 
architecture for these tools has remained largely unchanged for about two decades, limiting
learning capabilities. Here, we set out to improve the cross-species annotation of
genes from DNA sequence alone with the help of deep learning. The goal is to eliminate 
the dependency on a closely related gene model while also improving the predictive
quality in general with a fundamentally new architecture.\\
\textbf{Results:} We present Helixer, a framework for the development and usage of 
a cross-species deep learning model that improves significantly on performance and
generalizability when compared to more traditional methods. We evaluate our 
approach by building a single model for the annotation of 186 animal genomes as 
well as one for 51 plants. Our predictions are shown to be much less sensitive to
the length of the genome than those of a current state-of-the-art tool. We also 
present two novel post-processing techniques that each worked to further strengthen 
our annotations and show in-depth results of an RNA-Seq based comparison of our 
predictions.\\
\textbf{Availability:} The source code of this work is available at 
\url{https://github.com/weberlab-hhu/Helixer} under the GNU General 
Public License v3.0.\\
\textbf{Contact:} \href{alisandra.denton@hhu.de}{alisandra.denton@hhu.de}\\  
\textbf{Supplementary information:} Supplementary data are available at 
\textit{Bioinformatics} online.
}

@@

For the purpose of gene annotation, RNNs have already shown promising results. 
\citep{choudharypredicting} carries out preliminary explorations on the potential 
of BLSTMs for cross species gene prediction and trains his model on human genes to 
test it later on two more species. DeepAnnotator \citep{amin2018deepannotator} uses 
bidirectional LSTMs (BLSTM) for gene finding in prokaryotes. Gene prediction in 
prokaryotes is considered more amenable than in eukaryotes, as genes in prokaryotes 
are proportionately more frequent in the genome, feature simpler control structures 
and do not use splicing \citep{wang2004brief}. DanQ \citep{quang2016danq} proposes 
the use of a BLSTM after a convolutional neural net (CNN) to find detailed motifs in 
the human genome. DeePromoter \citep{oubounyt2019deepromoter} trains a similar 
architecture for the recognition of promoter regions. Recently, several groups 
\citep{jaganathan2019predicting, wang2019splicefinder} successfully used CNNs to 
find splicing sites.

In this work we present Helixer, a novel prototype software for training and 
utilizing a general purpose DNN for the ab-initio cross-species gene annotation of 
large eukaryotic genomes using only DNA sequence as input. Our model is trained to 
differentiate between 4 regions: Intergenic, Untranslated (UTR), Coding (CDS) and 
Intron. We demonstrate the effectiveness of this approach by training a single model 
for the annotation of a large set of genomes from the kingdoms metazoa and 
viridiplantae, respectively, which we will call {\it animals} and {\it plants} from 
now on. We worked with the full data of 192 animal genomes and 60 plant genomes. 
Both the ability to generalize across genomes as well as the scope of the evaluation 
greatly surpasses all other work in this field. The source code and all input data 
is publicly available.

\begin{methods}
\section{Methods}
\subsection{Datasets}
The foundation of our work are 192 animal and 60 plant genomes. The data of each 
genome consists of the latest publicly available genomic assembly in form of a 
FASTA file and the latest annotation in the GFF format. One genome for each species of animal 
in EnsemblMetazoa 45 (\cite{howe2020ensembl}; Table S\ref{suptab:downloads_animals}) 
as well as all non-embargoed plant species from the JGI Phytosome 13 database 
(\cite{goodstein2012phytozome}; Table S\ref{suptab:downloads_plants}); The exact 
genomes are listed in the aforementioned tables.
Data was downloaded on the 15/10/2019 and 29/03/2019 for animals and plants respectively. 
Both data groups were used separately during training and evaluation.

@@

For the selection of training genomes we had to balance multiple conflicting 
trade-offs. On the one hand, we want to train with as much data as possible, but 
on the other
not all data has sufficient quality to enable a powerful generalization. It is 
also very desirable to have diverse training genomes with a broad phylogenetic spread, 
a variety of genome sizes and average gene lengths as the model is increasingly unlikely to generalize 
well beyond the borders given by the training data. However, it also may be difficult 
for the model learn and generalize if the genomic patterns inside the data are 
too different from each other. Practically, it was important for us to limit the size of
our training genomes set to be small enough that enough that we could get experimental 
results within a couple of days and thereby be able to test many different data and 
model configurations.

@@

We call the genomes that were ultimately not used during training from the original 
split the {\it validation genomes}. While we did split off a group of test genomes 
at the beginning of our work, we ended up only reporting on the generalization 
performance on the combined set of validation and test genomes as the validation 
genomes are by far the largest set. We also did not find a noticeable difference in 
performance between those two groups.

@@

During data generation, we query for the transcript with the longest protein of each gene and 
disregard FASTA sequences that have no functional annotation as it is ambiguous
whether such sequences contain no genes or were simply not annotated in the reference. GeenuFF checks the 
genomic annotations for potential errors during import and is able to mark those 
areas. We use this information to effectively blank out those bases during training 
by using the sample weights described in Table \ref{tab:arrays}. The vast majority 
of masked bases lie in the intergenic region as the most prevalent error is a 
missing UTR and GeenuFF marks a potentially large intergenic region for it. Table 
\ref{tab:statistics_all} shows statistics about the masking.

For the training itself, we divide each continuous genomic sequence into 20,000 bp 
long subsequences, for which one-hot vectors of the base pairs and annotations 
are generated and respectively used as input and label for the neural network 
together with the sample weights. We append zero padding if the subsequences  
are shorter than 20,000 bp. 
If a subsequence is fully marked as erroneous, we exclude it from all analyses.

\begin{table}[!htb]
\processtable{Data group statistics for all data\label{tab:statistics_all}} {
\begin{tabular}{@{}lll@{}}
\toprule & Animals & Plants\\
\midrule
Average genome size in Gbp& 2.936 (+- 1.562) & 0.787 (+- 0.995) \\
Average gene length & 31,223 (+- 13,974)& 3,368 (+- 1,510)\\
Geenuff error rate & 0.311 (+- 0.129) & 0.351 (+- 0.249) \\
Fraction of class Intergenic  & 0.777 (+- 0.042) & 0.799 (+- 0.089) \\
Fraction of class UTR & 0.006 (+- 0.006) & 0.017 (+- 0.016) \\
Fraction of class CDS & 0.016 (+- 0.014) & 0.085 (+- 0.071) \\
Fraction of class Intron & 0.201 (+- 0.035) & 0.099 (+- 0.057) \\
\botrule
\end{tabular}}{{\it Note:} Values are averages of the individual values of each 
genome in a data group. All statistics except the average gene lengths exclude
FASTA sequences without a gene and any 20,000 bp subsequences that were masked as completely 
erroneous. Each strand of DNA was counted separately. The gene length was 
determined by the length of the pre-mRNA of the longest protein at each
loci. Brackets show the standard deviation.}
\end{table}

@@

\subsection{Model Architecture and Training}
\label{sec:model}
We use a 4-layer deep stacked BLSTM network with 256 units per layer and layer 
normalization \citep{ba2016layer} between each BLSTM layer to produce the 
predictions in the form of a base pair wise classification. The model consists of 
circa 5.4 million parameters and was implemented with the deep learning library 
Keras (https://keras.io) on top of TensorFlow \citep{abadi2016tensorflow}. We 
tested multiple different architectures before arriving at this model configuration, 
including convolutional neural networks (CNN) and hybrid architectures. We also use 
class weights as the class frequencies are both unbalanced and 
vary greatly between animals and plants. 
For example, the trade-off between the number of intergenic and intronic bases is 
quite different in both groups (see Table \ref{tab:statistics_all}).

@@

\subsection{Evaluation against independent RNAseq data}
\label{sec:rnaseq}
The reference annotations were created with existing tools and largely with a pipeline incorporating
{\it de novo} gene predictions with RNAseq and homology data. As the references, like any
data, are expected to contain errors we chose to use RNAseq data for an independent evaluation.
In plants there is the additional concern that the references may share biases with
AUGUSTUS or HMMs as AUGUSTUS was used as the {\it de novo} gene caller for the reference of
many plant species % (e.g. {\it Arabidopsis lyrata}, \cite{rawat2015improving}; 
%{\it Asparagus officinalis}, 
%\cite{harkess2017asparagus}; and {\it Chenopodium quinoa}, \cite{jarvis2017genome})  
and other HMM-based tools for many more. %(e.g. {\it Citrus clementina}, \cite{wu2018genomics}; and
%{\it Capsella grandiflora}, \cite{slotte2013capsella}). 
The Ensemble animal dataset by and large used the Ensembl annotation pipeline, which primarily
uses extrinsic data but never-the-less incorporates Genscan \citep{burge1997prediction} predictions.

@@

For each of these six species the following search was performed on Sequence Read Archive \citep{leinonen2010sequence}
`(("<species name>"[Organism] OR <species name>[All Fields]) AND stranded[All Fields]) 
AND ("biomol rna"[Properties] AND "library layout paired"[Properties])'. In so far as 
more than 50 samples were identified and every Nth sample was selected so that under 50
samples were chosen for further processing (see Table S\ref{suptab:rnaseq_samples}).
Each sample was prepped, mapped, and quality controlled in a pipeline using Trimmomatic 
\citep{bolger2014trimmomatic}, 
Hisat2 \citep{kim2019graph}, Samtools \citep{li2009sequence}, PicardTools 
(http://broadinstitute.github.io/picard/), 
FastQC 
(www.bioinformatics.babraham.ac.uk/projects/fastqc/),
and MultiQC (\cite{ewels2016multiqc}; see
Table S\ref{suptab:bioinfo_params} for details). The above was 
automated and the results visualized with the code here (https://github.com/weberlab-hhu/RNAsleek).
The samples were filtered to those with relatively high mapping rates, high mapping to
exonic relative to non-exonic regions, low 3' bias, normal FastQC and Trimmomatic stats, 
and a stranded mapping pattern (2nd read is sense strand). If more remained, seven
of the high quality samples were selected randomly.

Finally, the selected and mapped RNAseq samples (Table S\ref{suptab:rnaseq_samples}) were merged with Samtools 
and quantified to get the coverage (number of reads matching; i.e. cigar =, M or X),
and spliced coverage (number of reads with gap or splice; i.e. cigar N or D)
for every base pair in the genomes as implemented in Helixer's "training\_rnaseq.py"
script.

@@

Our models, however, tend to perform less well for the very smallest and largest 
genomes or species that are phylogenetically the furthest away from our training 
genomes. This is the case for both animals and plants and is visualized by the 
Figures S\ref{supfig:phylogeny_f1_animals} and S\ref{supfig:phylogeny_f1_plants}. 
We do not, for example, consistently predict very well on the 
algae nor on the non-avian reptiles. While one of the algae were included 
in our training genomes it accounted for a tiny proportion of the total training 
data and the non-avian reptiles were not included at all.

@@

\begin{figure*}[!h]
\label{fig:cov_example_main}
\centerline{\includegraphics[width=\textwidth]{images/cov_examples/cov_example_main}}
\caption{Four example helixer predictions in the context of RNAseq data, the reference,
and AUGUSTUS' prediction for {\it M. esculenta}. The examples were chosen so
that {\bf a)} Helixer had high accuracy against the reference and the reference
was supported by the RNAseq data, {\bf b)} Helixer had high accuracy but the 
reference was not supported by the RNAseq data, {\bf c)} Helixer had low accuracy
and the reference was supported by the RNAseq data, and {\bf d)} helixer had low
accuracy but the reference was not supported by RNAseq data. Feasibility for visualization was
a major secondary consideration. Each subplot shows from top to
bottom i) the natural log of the coverage (``cov", solid) and spliced coverage 
(``sc", dotted) + 1, ii) the reference annotation in matrix form, iii) 
AUGUSTUS' predictions in matrix form, and iv) Helixer's predictions. The reference
and AUGUSTUS have either 0 (white) or 1 (black) for each base pair and category, while
Helixer emits a probability from 0-1 represented via gray-scale. ``Ntrn" stand
for intron, and ``IG" stands for intergenic.
}
\end{figure*}

To make an independent qualitative and quantitative evaluation of
performance we compared Helixer predictions, AUGUSTUS predictions and the
reference to RNAseq coverage data. RNAseq evaluation was performed for
three each animals and plants. RNAseq coverage provides support for an 
exonic annotation (UTR or CDS), spliced coverages provides support for an 
annotation of intron, and neither coverage nor spliced coverage is expected 
for intergenic annotations. Looking at selected subsequences (Figure \ref{fig:cov_example_main}) % and supplemental figures.
we identified cases where RNAseq supported a) both the annotations of the reference 
and Helixer, b) neither, c) the reference, but not Helixer,
and d) Helixer, but not the reference.

Helixer models do not yet have postprocessing
to make finalized single predictions, but instead output base-wise probabilities.
We see that the model exhibits higher uncertainty around transitions between annotation 
classes, for instance between UTR and CDS, or more dramatically between UTR and 
intergenic even where Helixer predictions closely match the reference and RNAseq data (a). 
Helixer's uncertainty around transitions from UTR to intergenic regions may relate to a
fundamentally harder problem (there is no conserved motif at the site as is observed
for splice sites and start/stop codons), lack of a precise one 
base pair biological site \citep{carninci2006genome,hon2013quantification}
or noise in the reference, which we observed relative to the RNAseq data 
(Figures S\ref{supfig:UTRs_animals}-\ref{supfig:UTRs_plants}). % supplemental UTR figures, 1/species

Helixer models also sometimes showed uncertainty for larger regions.
In some cases, where Helixer did not receive RNAseq support for its highest probability
annotation, it assigned a low but non-trivial probability to the RNAseq-supported 
exon/intron pattern (Figures \ref{fig:cov_example_main}b-c, S\ref{supfig:cov_example_02}c). However, in the extreme,
there are cases where the Helixer model exhibits substantial indecision or confusion 
and shifts gradually between classes with no single class receiving a high probability 
for extended stretches (Figures S\ref{supfig:cov_example_02}c-d). 
Notably, in one of the examples (Figure S\ref{supfig:cov_example_02}d)
the RNAseq shows evidence of alternative splicing; 
and in another (Figure S\ref{supfig:cov_example_02}c) Helixer's
prediction falls between that of the reference and of AUGUSTUS.

@@

Coverage and Spliced coverage were broken down by the confusion matrix of both
Helixer vs AUGUSTUS as well as Helixer vs the reference for all genomes (Figure \ref{fig:average_vs_augustus}). 
Where both tools agree on far left in the figure, coverage and spliced coverage closely matched expectations. 
Specifically for CDS:CDS and UTR:UTR most base pairs had some, and many
had moderate or high coverage. The same pattern was seen for intron:intron and spliced
coverage. Finally, intergenic:intergenic showed only a small fraction of base pairs with 
any of either coverage or spliced coverage. In all conflicts the amount of RNAseq support
fell between the cases where tools agreed, indicating that all options were at least
capable of finding weak-spots in the other annotations.

In cases of conflict, Helixer's predictions were substantially more consistent with
the RNAseq data than those of AUGUSTUS. Specifically, for base pairs where one tool predicted CDS and
the other intergenic or intron, there was more coverage when 
Helixer predicted CDS (Figure \ref{fig:average_vs_augustus}a). Similarly, for base pairs where one tool predicted intron and the other 
CDS or intergenic, there was more spliced coverage when Helixer predicted intron (Figure \ref{fig:average_vs_augustus}b). 
These patterns were consistent in direction but varied in magnitude in the individual species,
with the exception of {\it P. marinus} where Helixer and AUGUSTUS performed comparably
at differentiating between CDS and intron regions (Figure S\ref{supfig:each_sp_vs_augustus}).

RNAseq based comparison of Helixer and the reference was less clear cut.
Averaged across species, helixer's predictions received slightly more support
than the reference when differentiating CDS, UTR and introns from intergenic;
however helixer's predictions received slightly less support than the reference 
when differentiating CDS and UTR from introns (Figure S\ref{supfig:average_vs_ref}). Performance varies between individual
species, from {\it P. marinus}, where the reference receives more support
in every conflict, to {\it T. cacao}, where Helixer models receive equivalent
or more support in every conflict (Figure S\ref{supfig:each_sp_vs_ref}). Interestingly, these two
species were selected as examples where Helixer had poor Subgenic F1; 
for the former the RNAseq confirms relatively weak performance for Helixer, while the 
for the latter RNAseq rather indicates a sub-par reference.

@@

Finally, development of a post-processing method
to go from base pair wise predictions to integrated predictions for whole 
transcripts at each loci could both further improve performance and would
greatly increase real world applications.
\section*{Funding}
This work has been supported by the Deutsche Forschungsgemeinschaft (DFG, German Research Foundation) 
under Germany's Excellence Strategy--EXC-2048/1--Project ID: 390686111.\vspace*{-12pt}
\section*{Acknowledgements}
We are grateful for the computational support and GPU access provided by
the de.NBI cloud, and by both the Centre for Information and Media Technology (ZIM)
and the Institute of Quantitative and Theoretical Biology at Heinrich Heine University.
